A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-05-30 text/manuscript.Rmd
update manuscript and references, remove roman numeral classes

---
title: "TTmanuscript"
author: "Dennis Hazelett"
date: "4/18/2020"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Title: A molecular taxonomy of tumors independent of tissue-of-origin

#### Authors: Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett

## Abstract: 
Cancer is a disease that affects metabolism and other basic cellular processes common to all cells. In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying theories elude the field. To address this challenge we reduced omics mutation and expression profiling data to pathway memberships. We found that tumors can be classified according to pathway disruption in twelve distinct classes orthogonal to a tissue of origin (including histological subtypes), number of mutations and clinical stage. Moreover, known germline risk mutations such as BRCA1/2 are equally distributed among the classes. One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR. This class is neatly subdivided into PI3K/Akt and Ras branches. Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling. Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling. Intriguingly, we identify a novel class of mutations in proteasome component proteins that may be sufficient to cause compound defects across multiple signaling pathways in this class. A third class is typified by defects in metabolism, especially of RNA and protein biosynthesis. In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors. With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found. We discuss the implications for tumor biology and treatment.

## Introduction:

Recent advances in technology have made possible the study of cancer at unprecedented levels of detail and scale. A combination of array and next-generation sequencing technologies has made it possible to assay the germline variation and somatic mutations, expression profile, methylation status, copy number variation. To take advantage of these tools, several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) endeavored to sequence large number of tumors (order of thousands) and collect data from multiple assays that could be analyzed together, with an eye towards integration and a general understanding of the mechanisms of cancer. These efforts have consisted chiefly of recognizing and quantifying the importance of individual genes in cancer etiology as the product of positive selection processes, or classification of cancers relative to common patterns of aberration in the expression of those genes in various assays.

Some considerable progress has been made in analyzing these data. Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance. Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies have been aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin. The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin. In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method [@shenIntegrativeClusteringMultiple2009], the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018]. The authors of this study analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.

@@

In this study, we attempted to address this challenge by focusing on discrete molecular pathways as the unit of disruption in cancer. This operation has the virtue of stripping all cell-type-specific information from the mutation data. We use the resulting statistic to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of tissue of origin. Furthermore, tumors of varying cancer type and/or histological subtype may merely reflect the likely cell-of-origin that gave birth to them.

## Results:

### Classification of tumors by pathway disruption reveals 3 major groups

To study cancer pathways we obtained a set of 7,607 tumor samples (x donors) from GDC. We wanted to classify tumors by pathway disruption considered different types of data available from TCGA. We chose to start with somatic mutations in exome sequencing data because the affected target gene is known. Therefore, we focused on all missense, nonsense, frameshift, and splicing mutations. We selected 381 Reactome pathways (reactome.org) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories (e.g. “transcription factors”) or disease associations (e.g. “mutated in colon cancer”) and filtered our gene list on membership in these pathways. We then compiled a mutation matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations.

In order to classify tumors using this dataset, we used unsupervised clustering methods. Preliminary work with hierarchical clustering produced a set of 15 clusters (optimized with clusterProfiler(ref) was very promising, as the tumors appeared to cluster largely independent of tissue of origin (Supplementary Figure S1A). To determine whether our clusters were differentiated from random selections of tumors we plotted Kaplan Meier survival curves for each of the clusters first, and compared the results to randomized sets of tumors of equal-sized proportions (Figure S1B, C). In addition we fitted a Cox proportional hazards test model, accounting for age of diagnosis and cluster membership and found that 8 of the clusters were significantly different from a reference cluster (cluster 1), suggesting that cluster membership could in principle be rooted in biological differences affecting survival (Supplementary Table S1). These differences were small relative to differences in cancer by tissue of origin (Figure S1D).

We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP (arXiv:1802.03426). Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings (Figure 1A). We call these tentatively by roman numerals I-V. While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure 1A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method. Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14). To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only). While there is a subtle trend, mostly visible for groups I and V (violin plot, Figure 1B), mutation count does not appear to be the dominant feature. To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure 1C).

The 15 hierarchical clusters were largely free of any tissue-of-origin bias, so we suspected that our 5 groups would be similarly independent from tissue of origin. To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures 1D). Full projections for all 23 cancers are presented in Supplementary Figure S2. Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies. One hypothesis to explain the lack of bias by cancer type is that the groupings are representative of these molecular subtypes. To probe this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP. Perhaps surprisingly we did not detect any group biases for these or any other subtype annotations in GDC (see Supplementary Figure S3 for the full set of projections), suggesting that these subtypes are reflective of differences either in specialized cell-type-of-origin or other cancer risk mechanisms.

There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes. However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length. BRCA1/2 genes for example are key for DNA double-stranded break repair (ref) and confer risk for breast, prostate and ovarian cancers (refs). Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure 1E). We made similar projections for the MMR genes MSH2, MSH6, MLH, PMS1 and PMS2; BRIP1, RAD15, CHEK2 and APC. None of these genes exhibited any remarkable specificity with respect to group (except maybe APC??) (Supplementary Figure S4). To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the TET2 and TET3 genes, plus telomerase 1, 2, 3, and observed similar lack of segregation by group (Supplementary Figure S4).

### Determinants of Tumor Class

In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 381 pathways used to produce it (Supplementary Figure S5), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment. We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure 2A). Using this approach, we identified 6? pathways associated with class II, 34 pathways primarily enriched in group III tumors, and 10 pathways enriched in class IV tumors. Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively. Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure 2B). Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution. Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure 2C). Class IV had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, beta-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others. As groups III and IV were the most distinct of the 5, we chose to investigate these two groups further.

Group III contained interesting sub-patterns of enrichment, which we subdivided further into class IIIa and class IIIb sequentially from right following the global naming system. Common to classes IIIa and b were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as EGFR, IGF1R, Insulin receptor, and ERBB2 and 4. In addition to these we observed SCF-Kit, DAP12, and Tie 2 signaling pathways, altogether suggesting that Class III tumors share mutations in extracellular signaling pathways generally. Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment. We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure 3 or supplement figure?). We find….(please do this analysis)

Intriguingly, we noticed that the pathways enriched for class IIIa and b were related to RAS and PI3K/Akt signaling, respectively, which potentially represent the downstream components of the extracellular signaling pathways discussed above. Moreover, class IIIb is also enriched in several immune-associated pathways including for both innate immunity and T cell receptor signaling, suggesting an interesting correlation between the PI3K/Akt subgroup and potential loss (or hyperactivation) of immune function. To address this possibility, we isolated the PI3K/Akt, RAS and immune/TCR pathways and assessed whether the mutations tended to be biased toward likely inactivating or activating mutations ….

Class IV exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors between class III and IV on the UMAP projection that have pathway characteristics of both. The top ten distinguishing pathways of Class IV tumors were cellular response to hypoxia, regulation of mitotic cell cycle, assembly of the pre-replicative complex, metabolism of polyamines, degradation of beta-catenin by the destruction complex, tumor necrosis factor receptor 2 non-canonical NF-kB pathway, AUF1 (hnRNP D0) binds and destabilizes RNA, regulation of apoptosis and 3 pathways associated with aspects of hedgehog signaling. To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (Figure 4A?). Surprisingly, this class of tumors fell into two major categories. The first contains many mutations, in nearly all of the hallmark pathways of this cluster, and the second subgroup contains far fewer mutations, but typically at least one mutation of any one of a large number of proteasome genes that happen to be common to the other 9 pathways. Moreover, it is apparent that some of these proteasome genes have a typically tissue-of-origin (or cancer-specific) mutation signature, suggesting a) that proteasome mutations could potentially substitute for a handful of more targeted mutations to these other pathways and b) that this could represent a novel class of cancer-promoting mutations that is obscured by virtue of having an abundance of targets with tissue-specific expression.

To further investigate this latter hypothesis, we examined the expression of the proteasome genes in this cohort of tumors, relating expression to percent of mutated samples by cancer (Figure 4B). We found a significant positive association between number of mutated samples (I’m making this up now, you have to determine whether it’s true and correct the record) and mean expression in the cancers by tissue-of-origin, thus confirming our suspicion that the proteasomal genes in class IV represent a novel class of tissue-specific specific cancer-promoting genes with a potential common mechanism.

### Shared mutated genes/pathways

@@

### Determinants of Metastatic vs non-metastatic tumors

Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure 2) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure 1C). Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown). We reasoned that the clusters might be different enough to dilute cluster-specific enrichment.

We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (Table 1). A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster. For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see supplemental figure S5) but not in 9 and 10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, Table 1). This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure 2) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7. Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.

There were also a number of pathways that were enriched with lower specificity (among 4 or more clusters) which were found to be enriched in metastases. We identified 18 such pathways, and of these 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing”. Of these pathways, we identified mutations in Sumoylation for 100% of the metastatic samples of clusters 3, 6, and 7 (Table 1).

@@

### Correlation of microenvironment with tumor class

## Materials & Methods

>>>>

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: "Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett"
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    code_folding: hide
    theme: flatly
    smart: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
```

## Abstract
Cancer is a disease that affects metabolism and other basic cellular processes common to all cells. In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying theories elude the field. To address this challenge we reduced omics mutation and expression profiling data to pathway memberships. We found that tumors can be classified according to pathway disruption in twelve distinct classes orthogonal to a tissue of origin (including histological subtypes), number of mutations and clinical stage. Moreover, known germline risk mutations such as BRCA1/2 are equally distributed among the classes. One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR. This class is neatly subdivided into PI3K/Akt and Ras signalling branches. Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling. Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling. Intriguingly, we identify a novel class of mutations in proteasome component proteins that may cause defects across multiple key signaling pathways. Yet another class is typified by defects in metabolism, especially of RNA and protein biosynthesis. In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors. With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found in. We discuss implications for tumor biology and treatment.

## Introduction

Recent advances in technology have made the study of cancer possible at unprecedented levels of detail and scale. A combination of array and next-generation sequencing technologies has made it possible to assay the germline variation and somatic mutations, expression profile, methylation status, copy number variation. To take advantage of these tools, several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) sequenced large numbers of tumors and collected data from multiple assays that could be analyzed together, with the goal of integration and increased general understanding of the mechanisms of cancer. These efforts have consisted chiefly of recognizing and quantifying the importance of genes in cancer etiology as the result of selection processes, or classification of cancers relative to common patterns of aberration in the expression of those genes in various assays.

Some considerable progress has been made analyzing these data. Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance. Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies have aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin. The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin. In a second study involving 33 cancer types and a much greater number of tumors using a newer algorithm (iCluster) [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018]. The authors analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.

@@

In this study, we attempted to address this challenge by focusing on discrete molecular pathways as the unit of disruption in cancer. Our approach has the virtue of stripping all cell-type-specific information from the mutation data and generalizing gene-level information to non-tissue specific cell-biology phenotypes. We use the resulting statistic to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of tissue of origin.

## Results

### Classification of tumors by pathway

To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We wanted to classify tumors by pathway disruption considered different types of data available from TCGA. We chose to start with somatic mutations in exome sequencing data because the affected target gene is known. Therefore, we focused on all missense, nonsense, frameshift, and splicing mutations. We selected 377 Reactome pathways [reactome.org](https://reactome.org/) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on membership in these pathways. We then compiled a matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations.

```{r data-set-creation, eval=FALSE}

```

In order to classify tumors using this dataset, we used unsupervised clustering methods. Preliminary work with hierarchical clustering produced a set of 15 clusters (optimized with clusterProfiler[@clusterProfiler] was very promising, as the tumors appeared to cluster largely independent of tissue of origin (**Supplementary Figure S1 A**).

```{r tumor-clusters}

```

To determine whether our clusters were differentiated from random selections of tumors we plotted Kaplan Meier survival curves for each of the clusters first, and compared the results to randomized sets of tumors of equal-sized proportions (**Supplementary Figures S1B-C**). In addition we fitted a Cox proportional hazards test model, accounting for age of diagnosis and cluster membership and found that 8 of the clusters were significantly different from a reference cluster (cluster 1), suggesting that cluster membership could in principle be related to biological differences affecting survival (**Supplementary Table S1**). These differences were modest relative to differences in cancer by tissue of origin (**Supplementary Figure S1D**).

```{r preliminary-survival}

```

We noted intriguing similarities across different clusters, which motivated us to use the dimensional reduction analysis (DRA) tool UMAP [@McInnes2018;@art-of-tsne]. Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings. To represent the spatial distribution in a more principled manner, we used hdbscan[@hdbscan] which resulted in 12 distinct new clusters projected onto the UMAP (**Figure 1A**). Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5 tumors from the hierarchical method predominate in cluster 1 of the DRA) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14 of the hierarchical method predominate in cluster 8 of the DRA). To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only). While there is a subtle trend, mostly visible for groups I and V (violin plot, **Figure 1B**), mutation count does not appear to be the dominant feature to distinguish the tumors in different classes of the DRA. To test whether the 5 groups represent progression by stage (which are determined clinically by size, though different criteria exist for different cancers), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every DRA cluster, about equally (**Figure \@ref(fig:umap-global-trends)C**).

```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
set.seed(21)
plot(x=rnorm(1000), y=rnorm(1000))
```

![**Figure 1 Classification of tumors independent of tissue of origin.** _**A.** UMAP: dimensional reduction analysis and clustering of pathway disruption data, followed by kNN assigns tumors to 12 distinct clusters. **B.** Violin plots of total mutations by pathway cluster reveal modest differences in overall mutation rate, except for cluster 8 which contains hypermutated tumors. In these tumors most pathways $>90%$ have at least one gene with a mutation. **C.** Tumors shaded by pathological stage on the UMAP from panel A do not favor any particular cluster. **D.** Metastatic status projected onto the UMAP clusters from panel A. Metastatic tumors are equally distributed among the clusters. **E.** Exemplars of 4 cancer types projected onto the UMAP clusters. Some biases are evident, but all cancers are found in every UMAP cluster. **F.** BRCA mutations, which are thought to increase genome instability, are found in every UMAP cluster._](../figures/fig1A_complete.png)

To determine whether the 12 DRA clusters are independent from tissue of origin, we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the original UMAP (**Figures \@ref(fig:umap-global-trends)D**). While some modest bias is evident in some cancers, No tissue specific cancer exclusively found in one cluster. Full projections for all 23 cancers are presented in **Supplementary Figure S2**. Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies. One hypothesis to explain the lack of bias by cancer type in these clusters is that the groupings are representative of these molecular subtypes. To test this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP. Perhaps surprisingly we did not observe any exclusive segregation by pathway for these subtype annotations in GDC (**Figure 1E**). We also projected histological subtype data for the remaining cancers with at best modest biases evident (see **Supplementary Figure S3** for the full set of projections), suggesting that these subtypes are as diverse with respect to our pathway disruption analysis as the parent cancer types.

There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes. However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to involve loss of heterozygosity [@brca-loh], so we projected the somatic mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe segregation of these mutations into specific clusters (**Figure \@ref(fig:umap-global-trends)E**). We made similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_, _PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes except for _APC_ exhibited any remarkable specificity with respect to cluster assignment (**Supplementary Figure S4**). To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and observed similar lack of segregation by cluster (**Supplementary Figure S4**).

### Determinants of Class

In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 377 pathways used to produce it (**Supplementary Figure S5**), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment. We visualized the enrichment as a heatmap (**Figure \@ref(fig:group-pathway-enrichment)A**). Using this approach, we identified 6? pathways associated with class II, 34 pathways primarily enriched in group III tumors, and 10 pathways enriched in class IV tumors. Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively. Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (**Figure \@ref(fig:group-pathway-enrichment)B**). Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution. Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (**Figure \@ref(fig:group-pathway-enrichment)C**). Cluster group 6-7 had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, $\beta$-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most distinct of the 12, we chose to investigate these two groups further.

```{r group-pathway-enrichment, fig.cap="Placeholder for figure 2", eval=FALSE}
set.seed(28)
plot(density(rnorm(100)))
```

![**Figure 2. Pathway-specific enrichment within clusters.** _Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns). The annotation at left indicates the Reactome pathway group at the second level (check this)._](../figures/fig2B_short.png)

Common to cluster groups 9-10 and 11-12 were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as _EGFR_, _IGF1R_, _Insulin receptor_, and _ERBB2_ and _-4_. In addition to these we observed _SCF-Kit_, _DAP12_, and _Tie 2_ signaling pathways, altogether suggesting that cluster group 9-12 tumors share mutations in extracellular signaling pathways generally.

```{r redundant-genes, fig.cap="Placeholder for figure 3", results=FALSE}
set.seed(29)
hist(rnorm(100))
```

Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment. Upon closer examination the split between cluster groups 9+10 and 11+12 is largely driven by mutations in _PI3K_ and its homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary Figure S4**). Furthermore, clusters 9 and 11 are distinguished from 10 and 12 by general lack of P53 mutations (compare **Figure 1A** with the _TP53_ panel of **Supplementary Figure S4**).

In contrast to cluster group 9-12, the group represented by clusters 6 & 7 exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors spatially distributed between cluster groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics of both. The top ten distinguishing pathways of class 6/7 tumors were cellular response to hypoxia, regulation of mitotic cell cycle, assembly of the pre-replicative complex, metabolism of polyamines, degradation of $\beta$-catenin by the destruction complex, tumor necrosis factor receptor 2 non-canonical _NF-$\kappa$B_ pathway, _AUF1_ (hnRNP D0) binds and destabilizes RNA, regulation of apoptosis and 3 pathways associated with aspects of hedgehog signaling. To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (Figure 4A?). Surprisingly, this class of tumors was not predominated by 1 or more highly mutated genes across all cancers. Instead, it appears that there is a whole class of genes in the proteosomal pathway whose activity contributes to all the aforementioned pathways, and the majority of tumors in these classes have a single mutation in one of these genes. Nonetheless, a significant fraction (~1/4 <-- need numbers!) do not have proteosomal mutations but instead harbor mutations in genes representing a majority of the same pathways, suggesting that most if not all of the pathways are individually important and that proteosomal mutations may be sufficient to drive the net defect of tumors in these clusters. We annotated the tumors in this class according to tissue of origin to determine whether proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and indeed we could observe some clustering. These data could explain how tissue-specific gene mutation signatures might result in general cancer phenotypes by converging at the pathway level.

@@

### Metastatic _vs._ non-metastatic tumors

Following the same logic we used to investigate cluster-specific enrichment of pathways (**Figure \@ref(fig:group-pathway-enrichment)**) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (**Figure \@ref(fig:umap-global-trends)C**). Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown). We reasoned that the clusters might be different enough to dilute cluster-specific enrichment.

We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (**Table 1**). A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster. For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see **Supplementary figure S5**) but not in 9 and 10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, **Table 1**). This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (**Figure \@ref(fig:group-pathway-enrichment)**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7. Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.

There were also a number of pathways that were enriched with lower specificity (among 4 or more clusters) which were found to be enriched in metastases. We identified 18 such pathways, and of these 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing”. Of these pathways, we identified mutations in Sumoylation for 100% of the metastatic samples of clusters 3, 6, and 7 (**Table 1**).

@@

### Survival Analysis

In order to understand whether our cluster identifications add any information to the prognositic value embedded in cancer type, we tested models of patient survival that accounts for cancer type and age at diagnosis. Initial models did not properly account for the trajectory of patient survival independent of age of diagnosis but showed some promise in distinguishing among classes for one or two of the deadliest cancer types. Therefore, we used mortality data from the CDC to accurately model the normal trajectory of patient survival independent of disease status, including gender as a factor. The resulting model is shown as a comparison with the census data in **Figure 6A.**
